---
title: "snRNA-seq Interactive Analysis — Fetal–Pediatric Lung"
author: "Spence Lab"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
    number_sections: true
    code_folding: show
editor_options:
  chunk_output_type: inline
---

<!-- ═══════════════════════════════════════════════════════════════════════════
     HOW TO USE THIS FILE
     ───────────────────
     1. Edit Section 0 (the CONFIG chunk below). Set all paths, checkpoints,
        and parameters there — nowhere else.
     2. Run chunks one at a time in RStudio (Ctrl/Cmd + Shift + Enter).
     3. Checkpoints: set CKPT_* to NULL to compute from scratch, or provide
        a file path to skip heavy steps and load a saved object instead.
     4. Save paths: set SAVE_* to NULL to skip saving, or provide a path.
     ═══════════════════════════════════════════════════════════════════════ -->

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(tidytext)
library(ggplot2)
library(patchwork)
library(Seurat)
library(SeuratWrappers)
library(sctransform)
library(slingshot)
library(SingleCellExperiment)
library(tradeSeq)
library(BiocParallel)
library(RColorBrewer)
library(viridis)
library(ComplexHeatmap)
library(ggridges)
library(ggrepel)
library(ggnewscale)
library(cowplot)
library(circlize)
library(presto)
library(writexl)
library(CellChat)
library(NMF)
library(ggalluvial)
library(R.utils)
```

---

# 0 · Configuration {.tabset}

> **Edit everything in this chunk, then run it first.**

```{r CONFIG}
# ╔══════════════════════════════════════════════════════════════════════════╗
# ║  USER CONFIG — edit here, run once, then step through the sections      ║
# ╚══════════════════════════════════════════════════════════════════════════╝

set.seed(123)
options(future.globals.maxSize = 40 * 1024^3)   # 40 GB; increase if needed


# ── Input data paths ──────────────────────────────────────────────────────────
PATH_fetal_raw     <- "PATH/TO/TF_fetal_nuc.rds"
PATH_pediatric_raw <- "PATH/TO/TF_ped_nuc.rds"


# ── Checkpoint paths ──────────────────────────────────────────────────────────
# Set to NULL → compute from scratch.
# Set to a file path → load that file and skip the heavy computation.

CKPT_fetal_integrated   <- NULL   # e.g. "outputs/rds/fetal_integrated.rds"
CKPT_ped_integrated     <- NULL   # e.g. "outputs/rds/ped_integrated.rds"
CKPT_joint_integrated   <- NULL   # e.g. "outputs/rds/joint_integrated.rds"
CKPT_epi_pre            <- NULL   # e.g. "outputs/rds/epi_pre.rds"
CKPT_epi_post           <- NULL   # e.g. "outputs/rds/epi_post.rds"
CKPT_alve_pre           <- NULL   # e.g. "outputs/rds/pre_alve.rds"
CKPT_alve_post          <- NULL   # e.g. "outputs/rds/post_alve.rds"
CKPT_sds_pre            <- NULL   # Slingshot SlingshotDataSet (pre)
CKPT_sds_post           <- NULL   # Slingshot SlingshotDataSet (post)
CKPT_sce_pre            <- NULL   # tradeSeq SCE (pre)
CKPT_sce_post           <- NULL   # tradeSeq SCE (post)
CKPT_cellchat_pre       <- NULL   # CellChat object (pre)
CKPT_cellchat_post      <- NULL   # CellChat object (post)


# ── Save paths ────────────────────────────────────────────────────────────────
# Set to NULL to skip saving. Set to a file path to save after computing.

SAVE_fetal_integrated   <- NULL   # e.g. "outputs/rds/fetal_integrated.rds"
SAVE_ped_integrated     <- NULL
SAVE_joint_integrated   <- NULL
SAVE_epi_pre            <- NULL
SAVE_epi_post           <- NULL
SAVE_alve_pre           <- NULL
SAVE_alve_post          <- NULL
SAVE_sds_pre            <- NULL
SAVE_sds_post           <- NULL
SAVE_sce_pre            <- NULL
SAVE_sce_post           <- NULL
SAVE_cellchat_pre       <- NULL
SAVE_cellchat_post      <- NULL


# ── Integration parameters ────────────────────────────────────────────────────
DIMS              <- 30
RESOLUTION        <- 0.5
VARS_TO_REGRESS   <- "percent.mt"
PED_DOWNSAMPLE_N  <- 27411   # pediatric cells to downsample to (match fetal)


# ── Contaminant gene filters (applied when subsetting epithelium) ─────────────
# Names = gene, values = expression threshold (cells above threshold excluded)
CONTAMINANT_GENES <- c(
  CLDN5  = 0.1,
  VWF    = 0.1,
  PTPRC  = 0.1,
  COL1A2 = 0.1,
  PECAM1 = 0.1,
  PDGFRA = 0.1
)


# ── Cell type labels to keep in the alveolar subset ───────────────────────────
ALVEOLAR_LABELS <- c(
  "AT1_1","AT1_2","AT1_3","AT1_4","AT1_5","AT1_6",
  "AT2_1","AT2_2","AT2_3","AT2_4","AT2_6","AT2_7",
  "Bud Tip",
  "Transitional","Transitional_1","Transitional_2",
  "Transitional_3","Transitional_4",
  "Transitional/Bud Tip Adjacent"
)


# ── Trajectory parameters ─────────────────────────────────────────────────────
TRAJ_START        <- "Bud Tip"    # starting cluster for Slingshot
TRAJ_DIST         <- "mnn"        # distance method: "mnn" or "slingshot"
TRAJ_DIST_2       <- "slingshot"  # secondary distance for comparison run
TRAJ_APPROX_PTS   <- 150
TRAJ_THRESH       <- 0.01
TRAJ_STRETCH      <- 0.8
TRAJ_SHRINK       <- 0.99
TRAJ_CLUSTER_COL  <- "cell_type2" # metadata column with cell type labels

# Lineage subsets to keep from each Slingshot run (NULL = keep all)
BT_MNN_LINEAGES   <- c(1, 3, 4)
T2_MNN_LINEAGE    <- 4
BT_SLING_LINEAGES <- c(2, 3)
T1_SLING_LINEAGE  <- 3


# ── tradeSeq parameters ───────────────────────────────────────────────────────
TRADESEQ_NKNOTS   <- 8
TRADESEQ_CORES    <- 10
TRADESEQ_GENES    <- NULL   # NULL = use VariableFeatures(); or provide a character vector

# Genes to visualize smoothers for interactively
SMOOTHER_GENES    <- c("CFTR", "SFTPC", "AGER", "SOX9")

# Pseudotime ridge: which cell types to show per lineage
RIDGE_KEEP <- list(
  Lineage1 = c("Bud Tip","AT2_1","AT2_2","Transitional_1","AT1_1","AT1_2"),
  Lineage2 = c("Bud Tip","AT2_1","AT2_2","Transitional_2"),
  Lineage3 = c("Bud Tip","AT2_1","AT2_2","AT2_4"),
  Lineage4 = c("Bud Tip","AT2_3")
)

# Heatmap: cell type order per lineage
HT_LINEAGE_ORDER <- list(
  Lineage1 = c("Bud Tip","AT2_2","Transitional","AT1_5","AT1_4","AT1_1","AT1_2"),
  Lineage2 = c("Bud Tip","AT2_2","Transitional","AT1_5","AT1_4","AT1_3"),
  Lineage3 = c("Bud Tip","AT2_2","AT2_1","AT2_3"),
  Lineage4 = c("Bud Tip","AT2_2","AT2_1","AT2_4")
)


# ── CellChat parameters ───────────────────────────────────────────────────────
CC_LABEL_COL        <- "cell_type2"  # metadata column for cell type labels
CC_SAMPLE_COL       <- "batch"
CC_MIN_CELLS        <- 10
CC_PATHWAYS         <- c("BMP","TGFb","FGF")  # pathways to visualize
CC_N_OUT_PATTERNS   <- 6
CC_N_IN_PATTERNS    <- 2
CC_SOURCES          <- c("Bud Tip","Transitional")  # bubble plot sources
CC_TARGETS          <- c("AT2_1","AT2_2","AT2_3","AT2_4")  # bubble plot targets
CC_CLUSTER_ORDER    <- c("Bud Tip","Transitional",
                          "AT2_1","AT2_2","AT2_3","AT2_4",
                          "AT1_1","AT1_2","AT1_3","AT1_4","AT1_5")


# ── Marker gene sets for FeaturePlots ─────────────────────────────────────────
MARKER_SETS <- list(
  AT2      = c("SFTPC","SFTPB","SFTPA1","SFTPA2","LAMP3","NAPSA"),
  AT1      = c("AGER","CAV1","EMP2","PDPN"),
  BudTip   = c("SOX9","TESC","CPM","ETV5","NPC2"),
  LAP      = c("SCGB3A2","CP","CFTR"),
  Ciliated = c("FOXJ1","CAPS","SNTN"),
  Endo     = c("CD34","PECAM1","VWF"),
  Alve     = c("SFTPC","SFTPB","LAMP3","NAPSA","SOX9","TESC","AGER","PDPN")
)

DOTPLOT_GENES <- c(
  "SOX9","TESC","CPM","ETV5","NPC2",
  "SFTPC","SFTPB","SFTPA1","SFTPA2","LAMP3","NAPSA",
  "AGER","CAV1","EMP2","PDPN",
  "SCGB3A2","CP","CFTR"
)


# ── Cluster annotation maps ───────────────────────────────────────────────────
# !! Update these when you re-cluster with different parameters !!

ANNOT_FETAL_LINEAGE <- c(
  "0"="Mesenchyme","1"="Mesenchyme","2"="Mesenchyme","3"="Mesenchyme",
  "4"="Epithelium","5"="Epithelium","6"="Epithelium","7"="Mesenchyme",
  "8"="Endothelium","9"="Mesenchyme","10"="Mesenchyme","11"="Immune",
  "12"="Epithelium","13"="Endothelium","14"="Immune","15"="Endothelium",
  "16"="Mesenchyme","17"="Epithelium","18"="Immune","19"="Endothelium"
)

ANNOT_PED_LINEAGE <- c(
  "0"="Epithelium","1"="Epithelium","2"="Epithelium","3"="Epithelium",
  "4"="Epithelium","5"="Mesenchyme","6"="Immune","7"="Epithelium",
  "8"="Endothelium","9"="Epithelium","10"="Immune","11"="Epithelium",
  "12"="Epithelium","13"="Immune","14"="Epithelium","15"="Immune",
  "16"="Endothelium","17"="Epithelium","18"="Endothelium","19"="Epithelium",
  "20"="Epithelium","21"="Epithelium","22"="Mesenchyme","23"="Mesenchyme",
  "24"="Epithelium","25"="Mesenchyme","26"="Mesenchyme","27"="Epithelium",
  "28"="Epithelium"
)

ANNOT_JOINT_CELLTYPE <- c(
  "0"="Fibroblast_1","1"="Fibroblast_2","2"="AT2_1","3"="AT1_1","4"="AT1_2",
  "5"="AT2_2","6"="Fibroblast_3","7"="Macrophage_1","8"="LAP","9"="Multiciliated",
  "10"="AT2_3","11"="Endothelial_1","12"="Transitional AT2","13"="Fibroblast_4",
  "14"="T-cell","15"="Low-quality/Dying/doublets","16"="Macrophage_2",
  "17"="Endothelial_2 (Capillary Aerocyte)","18"="Endothelial_3 (Lymphatic)",
  "19"="Monocytes","20"="Endothelial_4 (Pericyte)","21"="AT1_3","22"="Fibroblast_4",
  "23"="Basal","24"="Neutrophils","25"="B-cell","26"="Neuroendocrine",
  "27"="Unknown","28"="Endothelial_5"
)

ANNOT_JOINT_LINEAGE <- c(
  "0"="Mesenchyme","1"="Epithelium","2"="Epithelium","3"="Epithelium",
  "4"="Epithelium","5"="Mesenchyme","6"="Epithelium","7"="Immune",
  "8"="Mesenchyme","9"="Endothelium","10"="Epithelium","11"="Mesenchyme",
  "12"="Epithelium","13"="Immune","14"="Epithelium","15"="Epithelium",
  "16"="Epithelium","17"="Endothelium","18"="Immune","19"="Endothelium",
  "20"="Mesenchyme","21"="Endothelium","22"="Epithelium","23"="Immune",
  "24"="Immune","25"="Epithelium","26"="Epithelium"
)

ANNOT_EPI_POST_CELLTYPE <- c(
  "0"="AT2_1","1"="AT1_1","2"="LAP_1","3"="AT1_2","4"="AT2_2",
  "5"="Transitional_1","6"="Multiciliated","7"="AT2_3","8"="LAP_2",
  "9"="Transitional_2","10"="Endothelial_1","11"="Endothelial_2",
  "12"="AT1_3","13"="Basal","14"="Macrophages","15"="Neuroendocrine"
)

# Pre-alveolar subset annotation (16 clusters at res 0.6)
ANNOT_PRE_ALVE_CELLTYPE2 <- c(
  "0"="AT2_1","1"="Bud Tip","2"="AT1_1","3"="AT1_2","4"="AT1_3",
  "5"="AT2_2","6"="AT2_3","7"="AT2_4","8"="Transitional_1",
  "9"="Transitional_2","10"="Transitional_3","11"="Transitional_4",
  "12"="AT1_5","13"="AT2_6","14"="AT2_7","15"="AT1_6"
)

# Post-alveolar subset annotation (9 clusters)
ANNOT_POST_ALVE_CELLTYPE2 <- c(
  "0"="AT2_1","1"="AT1_1","2"="AT2_2","3"="Transitional_1",
  "4"="AT1_2","5"="Transitional_2","6"="AT2_4","7"="Bud Tip","8"="AT2_5"
)

# Additional celltype3 annotation for post-alve (collapsed Transitional/Bud Tip)
ANNOT_POST_ALVE_CELLTYPE3 <- c(
  "0"="AT2_1","1"="AT1_1","2"="AT2_2","3"="AT2_3","4"="AT1_2",
  "5"="Transitional","6"="Bud Tip","7"="AT1_5","8"="AT2_4",
  "9"="AT1_3","10"="AT1_4"
)

# ── Plot settings ─────────────────────────────────────────────────────────────
FIG_W     <- 12
FIG_H     <- 8
PT_SIZE   <- 0.5
RASTER    <- FALSE   # TRUE speeds up rendering for large datasets

# Sequential color gradient (used in FeaturePlots and pseudotime)
MY_COLORS <- scale_color_gradientn(colors = c(
  "lightgrey","#EDF8B1","#C7E9B4","#7FCDBB","#41B6C4",
  "#1D91C0","#225EA8","#253494","#081D58"
))
MY_COLORS_FN <- colorRampPalette(rev(c(
  "#EDF8B1","#C7E9B4","#7FCDBB","#41B6C4",
  "#1D91C0","#225EA8","#253494","#081D58"
)))

# Group colors
COL_FETAL <- "#6A3D9A"
COL_PED   <- "#F28E2B"

cat("✓ Config loaded.\n")
```

```{r helpers, include=FALSE}
# Internal helpers — no need to edit

scale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!is.finite(rng[1]) || !is.finite(rng[2]) || rng[2] == rng[1])
    return(rep(NA_real_, length(x)))
  (x - rng[1]) / (rng[2] - rng[1])
}

combined_pt_from_sds <- function(sds, cells) {
  pt <- slingPseudotime(sds)
  w  <- slingCurveWeights(sds)
  pt <- pt[cells, , drop = FALSE]
  w  <- w[cells,  , drop = FALSE]
  pt_scaled <- apply(pt, 2, scale01)
  w_sum     <- rowSums(w, na.rm = TRUE)
  pt_one    <- rowSums(pt_scaled * w, na.rm = TRUE) / w_sum
  pt_one[!is.finite(pt_one)] <- NA_real_
  pt_one
}

normalize_labels <- function(x) {
  x <- as.character(x); x <- trimws(x); x <- tolower(x)
  x <- gsub("[^a-z0-9]+", "_", x); x <- gsub("^_|_$", "", x); x
}

run_integration <- function(obj, dims = DIMS, resolution = RESOLUTION,
                             split_by = "batch", regress = VARS_TO_REGRESS) {
  DefaultAssay(obj) <- "RNA"
  obj <- JoinLayers(obj)
  obj[["RNA"]] <- split(obj[["RNA"]], f = obj@meta.data[[split_by]])
  obj <- SCTransform(obj, vars.to.regress = regress, verbose = FALSE)
  obj <- RunPCA(obj, dims = seq_len(dims), verbose = FALSE)
  obj <- IntegrateLayers(obj, method = RPCAIntegration,
                          normalization.method = "SCT", verbose = FALSE)
  obj[["RNA"]] <- JoinLayers(obj[["RNA"]])
  red <- if ("integrated.dr" %in% names(obj@reductions)) "integrated.dr" else "pca"
  obj <- FindNeighbors(obj, dims = seq_len(dims), reduction = red, verbose = FALSE)
  obj <- FindClusters(obj, resolution = resolution)
  obj <- RunUMAP(obj, dims = seq_len(dims), reduction = red)
  obj
}

load_or_compute <- function(ckpt, compute_fn, save_path = NULL, label = "") {
  if (!is.null(ckpt) && file.exists(ckpt)) {
    message("✓ [", label, "] Loading checkpoint: ", ckpt)
    readRDS(ckpt)
  } else {
    message("⟳ [", label, "] Computing ...")
    obj <- compute_fn()
    if (!is.null(save_path)) {
      dir.create(dirname(save_path), recursive = TRUE, showWarnings = FALSE)
      saveRDS(obj, save_path)
      message("  Saved → ", save_path)
    }
    obj
  }
}

build_contaminant_expr <- function(gene_thresholds) {
  paste(
    mapply(function(g, t) sprintf("%s < %g", g, t),
           names(gene_thresholds), gene_thresholds),
    collapse = " & "
  )
}

quick_dimplots <- function(obj, group_vars, prefix = "") {
  plots <- lapply(group_vars, function(gv) {
    if (!gv %in% colnames(obj@meta.data)) return(NULL)
    DimPlot(obj, group.by = gv, reduction = "umap",
            label = FALSE, raster = RASTER, pt.size = PT_SIZE) +
      theme_void() + ggtitle(paste0(prefix, gv))
  })
  plots <- Filter(Negate(is.null), plots)
  wrap_plots(plots, ncol = 2)
}

scale_rows <- function(x) {
  m <- rowMeans(x, na.rm = TRUE)
  s <- apply(x, 1, sd, na.rm = TRUE); s[s == 0 | is.na(s)] <- 1
  sweep(sweep(x, 1, m, "-"), 1, s, "/")
}
```

---

# 1 · Data Input

```{r load-raw}
TF_fetal_nuc     <- readRDS(PATH_fetal_raw)
TF_pediatric_nuc <- readRDS(PATH_pediatric_raw)

cat("Fetal:    ", ncol(TF_fetal_nuc), "cells ×", nrow(TF_fetal_nuc), "genes\n")
cat("Pediatric:", ncol(TF_pediatric_nuc), "cells ×", nrow(TF_pediatric_nuc), "genes\n")
cat("\nFetal metadata columns:\n"); print(colnames(TF_fetal_nuc@meta.data))
cat("\nPediatric metadata columns:\n"); print(colnames(TF_pediatric_nuc@meta.data))
cat("\nFetal cells per batch:\n"); print(table(TF_fetal_nuc$batch))
cat("\nPediatric cells per batch:\n"); print(table(TF_pediatric_nuc$batch))
```

---

# 2 · Individual Integration {.tabset}

## 2a · Fetal

```{r integrate-fetal}
TF_fetal_nuc <- load_or_compute(
  ckpt       = CKPT_fetal_integrated,
  save_path  = SAVE_fetal_integrated,
  label      = "fetal integration",
  compute_fn = function() run_integration(TF_fetal_nuc)
)
```

```{r plot-fetal, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_fetal_nuc,
               c("seurat_clusters","batch","sex","age","TF_annotation"),
               prefix = "Fetal – ")
```

## 2b · Pediatric

```{r integrate-pediatric}
# Downsample then integrate
TF_pediatric_nuc_subset <- load_or_compute(
  ckpt       = CKPT_ped_integrated,
  save_path  = SAVE_ped_integrated,
  label      = "pediatric integration",
  compute_fn = function() {
    set.seed(123)
    sub <- subset(TF_pediatric_nuc,
                  cells = sample(colnames(TF_pediatric_nuc), PED_DOWNSAMPLE_N))
    run_integration(sub)
  }
)
cat("Pediatric (downsampled):", ncol(TF_pediatric_nuc_subset), "cells\n")
```

```{r plot-pediatric, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_pediatric_nuc_subset,
               c("seurat_clusters","batch","sex","age",
                 "TF_annotation_lvl1","TF_annotation_lvl2","TF_annotation_lvl3"),
               prefix = "Pediatric – ")
```

---

# 3 · Lineage Annotation

```{r annotate-fetal}
TF_fetal_nuc@meta.data$lineage_lvl1 <-
  ANNOT_FETAL_LINEAGE[as.character(TF_fetal_nuc$seurat_clusters)]

DimPlot(TF_fetal_nuc, group.by = "lineage_lvl1",
        reduction = "umap", label = FALSE, raster = RASTER) +
  theme_void() + ggtitle("Fetal – Lineage")
```

```{r annotate-pediatric}
TF_pediatric_nuc_subset@meta.data$lineage_lvl1 <-
  ANNOT_PED_LINEAGE[as.character(TF_pediatric_nuc_subset$seurat_clusters)]

DimPlot(TF_pediatric_nuc_subset, group.by = "lineage_lvl1",
        reduction = "umap", label = FALSE, raster = RASTER) +
  theme_void() + ggtitle("Pediatric – Lineage")
```

---

# 4 · Joint Fetal–Pediatric Integration

```{r integrate-joint}
TF_fetal_pediatric <- load_or_compute(
  ckpt       = CKPT_joint_integrated,
  save_path  = SAVE_joint_integrated,
  label      = "joint integration",
  compute_fn = function() {
    merged <- merge(TF_fetal_nuc, TF_pediatric_nuc_subset)
    run_integration(merged)
  }
)
cat("Joint object:", ncol(TF_fetal_pediatric), "cells\n")
```

```{r annotate-joint}
TF_fetal_pediatric$cluster   <- Idents(TF_fetal_pediatric)
TF_fetal_pediatric$cell_type <- ANNOT_JOINT_CELLTYPE[as.character(TF_fetal_pediatric$cluster)]
TF_fetal_pediatric$lineage_lvl2 <- ANNOT_JOINT_LINEAGE[as.character(TF_fetal_pediatric$cluster)]
```

```{r plot-joint, fig.width=FIG_W*2, fig.height=FIG_H}
quick_dimplots(TF_fetal_pediatric,
               c("seurat_clusters","lineage_lvl2","cell_type","group","batch"),
               prefix = "Joint – ")
```

```{r joint-annotation-crosswalk}
# Dominant TF annotation per cluster — use this table to review and update
# ANNOT_JOINT_CELLTYPE above if needed.
TF_fetal_pediatric@meta.data %>%
  filter(!is.na(TF_annotation_lvl1)) %>%
  group_by(cluster, TF_annotation_lvl1) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(pct = n / sum(n) * 100) %>%
  slice_max(pct, n = 1, with_ties = FALSE) %>%
  arrange(as.numeric(as.character(cluster)))
```

---

# 5 · Epithelial Subsetting {.tabset}

## 5a · Pre-merge (subset → merge → integrate)

```{r epi-pre}
TF_fetal_pediatric_epi_pre <- load_or_compute(
  ckpt       = CKPT_epi_pre,
  save_path  = SAVE_epi_pre,
  label      = "epi pre-merge",
  compute_fn = function() {
    # Build contaminant filter expression
    cont_expr <- build_contaminant_expr(CONTAMINANT_GENES)
    fetal_filt <- sprintf("lineage_lvl1 == 'Epithelium' & %s", cont_expr)
    ped_filt   <- sprintf("lineage_lvl1 == 'Epithelium' & %s", cont_expr)

    fetal_epi <- subset(TF_fetal_nuc,
                         subset = eval(str2expression(fetal_filt)))
    ped_epi   <- subset(TF_pediatric_nuc_subset,
                         subset = eval(str2expression(ped_filt)))

    cat("Fetal epithelial cells:    ", ncol(fetal_epi), "\n")
    cat("Pediatric epithelial cells:", ncol(ped_epi),   "\n")

    merged <- merge(fetal_epi, ped_epi)
    DefaultAssay(merged) <- "RNA"
    run_integration(merged)
  }
)
cat("Pre-merge epi:", ncol(TF_fetal_pediatric_epi_pre), "cells\n")
```

```{r plot-epi-pre, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_fetal_pediatric_epi_pre,
               c("seurat_clusters","batch","sex","group","TF_annotation"),
               prefix = "Epi Pre – ")
```

## 5b · Post-merge (joint object → subset → integrate)

```{r epi-post}
TF_fetal_pediatric_epi_post <- load_or_compute(
  ckpt       = CKPT_epi_post,
  save_path  = SAVE_epi_post,
  label      = "epi post-merge",
  compute_fn = function() {
    # Fix: explicit assignment of contaminant-filtered subset (bug from original)
    cont_expr <- build_contaminant_expr(CONTAMINANT_GENES)
    filt_expr  <- sprintf("lineage_lvl2 == 'Epithelium' & %s", cont_expr)
    obj <- subset(TF_fetal_pediatric,
                   subset = eval(str2expression(filt_expr)))
    cat("Post-merge epithelial cells:", ncol(obj), "\n")
    run_integration(obj)
  }
)
cat("Post-merge epi:", ncol(TF_fetal_pediatric_epi_post), "cells\n")
```

```{r annotate-epi-post}
TF_fetal_pediatric_epi_post$cluster   <- Idents(TF_fetal_pediatric_epi_post)
TF_fetal_pediatric_epi_post$cell_type2 <-
  ANNOT_EPI_POST_CELLTYPE[as.character(TF_fetal_pediatric_epi_post$cluster)]
```

```{r plot-epi-post, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_fetal_pediatric_epi_post,
               c("seurat_clusters","cell_type2","batch","group","TF_annotation_lvl3"),
               prefix = "Epi Post – ")
```

## 5c · Epithelial markers and crosswalk

```{r epi-markers}
# Top markers per cluster to guide cell type assignment
res_epi       <- wilcoxauc(TF_fetal_pediatric_epi_post, seurat_assay = "RNA")
top_markers_epi <- top_markers(res_epi, n = 10)
print(top_markers_epi)
```

```{r epi-crosswalk}
TF_fetal_pediatric_epi_post@meta.data %>%
  filter(!is.na(TF_annotation_lvl3)) %>%
  group_by(cluster, TF_annotation_lvl3) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(pct = n / sum(n) * 100) %>%
  slice_max(pct, n = 1, with_ties = FALSE) %>%
  arrange(as.numeric(as.character(cluster)))
```

```{r epi-featureplot, fig.width=FIG_W, fig.height=8}
DefaultAssay(TF_fetal_pediatric_epi_post) <- "SCT"
FeaturePlot(TF_fetal_pediatric_epi_post,
            features = c("SOX9","TESC","NPC2","SFTPC","AGER","SCGB3A2"),
            ncol = 3, pt.size = 0.2, raster = RASTER) &
  MY_COLORS & theme_void()
```

---

# 6 · Alveolar Subsetting {.tabset}

## 6a · Pre-alveolar

```{r alve-pre}
TF_fetal_pediatric_epi_pre_alve <- load_or_compute(
  ckpt       = CKPT_alve_pre,
  save_path  = SAVE_alve_pre,
  label      = "pre-alveolar",
  compute_fn = function() {
    # Annotate pre-epi first
    TF_fetal_pediatric_epi_pre$cluster    <- Idents(TF_fetal_pediatric_epi_pre)
    TF_fetal_pediatric_epi_pre$cell_type2 <-
      ANNOT_PRE_ALVE_CELLTYPE2[as.character(TF_fetal_pediatric_epi_pre$seurat_clusters)]

    obj <- subset(TF_fetal_pediatric_epi_pre,
                   subset = cell_type2 %in% ALVEOLAR_LABELS)
    cat("Pre-alveolar cells:", ncol(obj), "\n")
    run_integration(obj)
  }
)

TF_fetal_pediatric_epi_pre_alve$cell_type2 <-
  ANNOT_PRE_ALVE_CELLTYPE2[as.character(TF_fetal_pediatric_epi_pre_alve$seurat_clusters)]

cat("Pre-alveolar:", ncol(TF_fetal_pediatric_epi_pre_alve), "cells\n")
```

```{r plot-alve-pre, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_fetal_pediatric_epi_pre_alve,
               c("cell_type2","group","batch"), prefix = "Pre-Alve – ")
```

## 6b · Post-alveolar

```{r alve-post}
TF_fetal_pediatric_epi_post_alve <- load_or_compute(
  ckpt       = CKPT_alve_post,
  save_path  = SAVE_alve_post,
  label      = "post-alveolar",
  compute_fn = function() {
    obj <- subset(TF_fetal_pediatric_epi_post,
                   subset = cell_type2 %in% ALVEOLAR_LABELS)
    cat("Post-alveolar cells:", ncol(obj), "\n")
    run_integration(obj)
  }
)

TF_fetal_pediatric_epi_post_alve$cell_type2 <-
  ANNOT_POST_ALVE_CELLTYPE2[as.character(TF_fetal_pediatric_epi_post_alve$seurat_clusters)]
TF_fetal_pediatric_epi_post_alve$cell_type3 <-
  ANNOT_POST_ALVE_CELLTYPE3[as.character(TF_fetal_pediatric_epi_post_alve$seurat_clusters)]

cat("Post-alveolar:", ncol(TF_fetal_pediatric_epi_post_alve), "cells\n")
```

```{r plot-alve-post, fig.width=FIG_W, fig.height=FIG_H}
quick_dimplots(TF_fetal_pediatric_epi_post_alve,
               c("cell_type2","cell_type3","group","batch"), prefix = "Post-Alve – ")
```

---

# 7 · Alveolar Visualization {.tabset}

## 7a · Feature plots

```{r featureplot-alve, fig.width=FIG_W, fig.height=10}
DefaultAssay(TF_fetal_pediatric_epi_post_alve) <- "SCT"
for (nm in names(MARKER_SETS)) {
  genes <- MARKER_SETS[[nm]]
  genes <- genes[genes %in% rownames(TF_fetal_pediatric_epi_post_alve)]
  if (length(genes) == 0) next
  p <- FeaturePlot(TF_fetal_pediatric_epi_post_alve,
                   features = genes, pt.size = 0.2,
                   raster = RASTER, combine = FALSE)
  p <- lapply(seq_along(p), function(i)
    p[[i]] + MY_COLORS + theme_void() + ggtitle(genes[i]) +
      theme(plot.title = element_text(hjust = 0.5, size = 12)))
  print(wrap_plots(p, ncol = 2) +
          plot_annotation(title = paste(nm, "Markers"),
                          theme = theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))))
}
```

## 7b · Age gradient UMAP

```{r age-gradient, fig.width=10, fig.height=5}
meta    <- TF_fetal_pediatric_epi_post_alve@meta.data %>% rownames_to_column("cell")
umap_df <- Embeddings(TF_fetal_pediatric_epi_post_alve, "umap") %>%
  as.data.frame() %>% rownames_to_column("cell")
plot_df <- left_join(umap_df, meta, by = "cell") %>%
  filter(group %in% c("fetal","pediatric"), !is.na(age_days))

ggplot() +
  geom_point(data = filter(plot_df, group == "fetal"),
             aes(umap_1, umap_2, color = age_days), size = PT_SIZE) +
  scale_color_gradientn(colors = brewer.pal(9,"Purples"),
                        name = "Fetal Age (days)",
                        limits = range(filter(plot_df, group=="fetal")$age_days, na.rm=TRUE),
                        na.value = "transparent") +
  new_scale_color() +
  geom_point(data = filter(plot_df, group == "pediatric"),
             aes(umap_1, umap_2, color = age_days), size = PT_SIZE) +
  scale_color_gradientn(colors = brewer.pal(9,"Oranges"),
                        name = "Pediatric Age (days)",
                        limits = range(filter(plot_df, group=="pediatric")$age_days, na.rm=TRUE),
                        na.value = "transparent") +
  theme_void() +
  theme(legend.text = element_text(size = 12),
        legend.title = element_text(size = 14, face = "bold"))
```

## 7c · Dot plot

```{r dotplot, fig.width=14, fig.height=6}
# Change cell_type2 below to cell_type3 if using the collapsed annotation
genes_use <- DOTPLOT_GENES[DOTPLOT_GENES %in% rownames(TF_fetal_pediatric_epi_pre_alve)]

dd <- FetchData(TF_fetal_pediatric_epi_pre_alve,
                vars = c("cell_type2", genes_use)) %>%
  pivot_longer(all_of(genes_use), names_to = "gene", values_to = "expr") %>%
  group_by(cell_type2, gene) %>%
  summarise(avg_expr = mean(expr, na.rm = TRUE),
            pct      = mean(expr > 0) * 100, .groups = "drop") %>%
  mutate(gene = factor(gene, levels = genes_use))

ggplot(dd, aes(x = gene, y = cell_type2, size = pct, color = avg_expr)) +
  geom_point() +
  scale_color_gradientn(colors = c("lightgrey","#EDF8B1","#C7E9B4","#7FCDBB",
                                    "#41B6C4","#1D91C0","#225EA8","#253494","#081D58")) +
  scale_size_area(max_size = 9) +
  labs(x = "Gene", y = "Cell type",
       size = "% expressing", color = "Avg expression") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        panel.grid  = element_blank())
```

## 7d · DEG heatmap

```{r deg-heatmap, fig.width=10, fig.height=20}
Idents(TF_fetal_pediatric_epi_post_alve) <- "cell_type3"

deg_list <- lapply(levels(Idents(TF_fetal_pediatric_epi_post_alve)), function(ct) {
  tryCatch(
    FindMarkers(TF_fetal_pediatric_epi_post_alve, ident.1 = ct,
                only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25),
    error = function(e) NULL
  )
})
names(deg_list) <- levels(Idents(TF_fetal_pediatric_epi_post_alve))

top_genes <- unique(unlist(lapply(deg_list, function(d) {
  if (is.null(d) || nrow(d) == 0) return(NULL)
  head(rownames(d[order(-d$avg_log2FC), ]), 10)
})))

if (length(top_genes) > 0) {
  expr_df <- FetchData(TF_fetal_pediatric_epi_post_alve,
                        vars = c(top_genes, "cell_type3")) %>%
    pivot_longer(-cell_type3, names_to = "gene", values_to = "expr") %>%
    group_by(cell_type3, gene) %>%
    summarise(avg = mean(expr, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = cell_type3, values_from = avg) %>%
    column_to_rownames("gene")

  scaled <- t(scale(t(as.matrix(expr_df))))
  Heatmap(scaled, name = "Z-score",
          cluster_rows = TRUE, cluster_columns = TRUE,
          col = viridis(100), show_row_names = TRUE,
          row_names_gp = gpar(fontsize = 6),
          column_names_gp = gpar(fontsize = 11, rot = 45))
}
```

---

# 8 · Slingshot Trajectory {.tabset}

## 8a · Pre-merge trajectory

```{r slingshot-pre}
# Re-cluster at res 0.6 for trajectory (finer clusters)
DefaultAssay(TF_fetal_pediatric_epi_pre_alve) <- "SCT"
red_pre <- if ("integrated.dr" %in% names(TF_fetal_pediatric_epi_pre_alve@reductions))
  "integrated.dr" else "pca"
TF_fetal_pediatric_epi_pre_alve <- FindNeighbors(
  TF_fetal_pediatric_epi_pre_alve, dims = seq_len(DIMS), reduction = red_pre, verbose = FALSE)
TF_fetal_pediatric_epi_pre_alve <- FindClusters(
  TF_fetal_pediatric_epi_pre_alve, resolution = 0.6)
TF_fetal_pediatric_epi_pre_alve <- RunUMAP(
  TF_fetal_pediatric_epi_pre_alve, dims = seq_len(DIMS), reduction = red_pre)
TF_fetal_pediatric_epi_pre_alve$cell_type2 <-
  ANNOT_PRE_ALVE_CELLTYPE2[as.character(TF_fetal_pediatric_epi_pre_alve$seurat_clusters)]

dimred_pre  <- TF_fetal_pediatric_epi_pre_alve@reductions$umap@cell.embeddings
clustering_pre <- TF_fetal_pediatric_epi_pre_alve$cell_type2
counts_pre  <- GetAssayData(TF_fetal_pediatric_epi_pre_alve, assay = "RNA", layer = "counts")

sds_pre <- load_or_compute(
  ckpt       = CKPT_sds_pre,
  save_path  = SAVE_sds_pre,
  label      = "slingshot pre",
  compute_fn = function() {
    lin <- getLineages(data = dimred_pre, clusterLabels = clustering_pre,
                       start.clus = TRAJ_START, dist.method = TRAJ_DIST)
    sds <- getCurves(lin, approx_points = TRAJ_APPROX_PTS, thresh = TRAJ_THRESH,
                     stretch = TRAJ_STRETCH, allow.breaks = FALSE, shrink = TRAJ_SHRINK)
    SlingshotDataSet(sds)
  }
)
cat("Pre lineages:", length(sds_pre@lineages), "\n")
```

```{r plot-slingshot-pre, fig.width=FIG_W, fig.height=FIG_H}
# Extract Seurat cluster colors
Idents(TF_fetal_pediatric_epi_pre_alve) <- "cell_type2"
p_col <- DimPlot(TF_fetal_pediatric_epi_pre_alve, group.by = "cell_type2")
pb    <- ggplot_build(p_col)
col_levels  <- levels(Idents(TF_fetal_pediatric_epi_pre_alve))
umap_colors <- setNames(unique(pb$data[[1]]$colour), col_levels)

plot(dimred_pre[,1:2], col = umap_colors[as.character(clustering_pre)],
     cex = 0.5, pch = 16, xlab = "", ylab = "", axes = FALSE)
lines(sds_pre, lwd = 4, col = adjustcolor("steelblue", 0.8))
for (cl in unique(as.character(clustering_pre))) {
  idx <- which(as.character(clustering_pre) == cl)
  if (length(idx) > 0)
    text(mean(dimred_pre[idx,1]), mean(dimred_pre[idx,2]), cl, font = 2, cex = 0.9)
}
```

```{r pseudotime-pre, fig.width=FIG_W, fig.height=FIG_H}
pt_all_pre  <- slingPseudotime(sds_pre)
pt_comb_pre <- combined_pt_from_sds(sds_pre, rownames(dimred_pre))
pt_comb_pre[is.na(pt_comb_pre)] <- 0

pal_pre  <- MY_COLORS_FN(200)
idx_pre  <- pmax(1, pmin(200, 1 + floor(scale01(pt_comb_pre) * 199)))
pt_cols  <- pal_pre[idx_pre]

plot(dimred_pre[,1:2], col = pt_cols, pch = 16, cex = 0.5, axes = FALSE,
     xlab = "", ylab = "", main = "Pre-alve: weighted pseudotime")
for (cl in unique(as.character(clustering_pre))) {
  idx <- which(as.character(clustering_pre) == cl)
  if (length(idx) > 0)
    text(mean(dimred_pre[idx,1]), mean(dimred_pre[idx,2]), cl, font = 2, cex = 0.9, col = "darkred")
}
```

```{r ridge-pre, fig.width=8, fig.height=12}
pt_long_pre <- as.data.frame(pt_all_pre) %>%
  mutate(cell_type = clustering_pre) %>%
  pivot_longer(colnames(pt_all_pre), names_to = "lineage", values_to = "pt") %>%
  filter(!is.na(pt)) %>%
  filter(mapply(function(ln, ct) isTRUE(ct %in% RIDGE_KEEP[[ln]]), lineage, cell_type))

meds_pre <- pt_long_pre %>%
  group_by(lineage, cell_type) %>% summarize(med = median(pt), .groups = "drop")

pt_long_pre %>%
  left_join(meds_pre, by = c("lineage","cell_type")) %>%
  mutate(cell_ord = reorder_within(cell_type, med, within = lineage)) %>%
  ggplot(aes(x = pt, y = cell_ord, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 2.2, rel_min_height = 0.01,
                                size = 0.2, color = "white") +
  scale_fill_gradientn(colors = MY_COLORS_FN(200), name = "Pseudotime") +
  scale_y_reordered() +
  facet_wrap(~lineage, ncol = 1, scales = "free_y") +
  labs(x = "Pseudotime", y = NULL, title = "Pre — pseudotime by cell type") +
  theme_ridges() + theme(panel.grid = element_blank())
```

## 8b · Post-merge trajectory

```{r slingshot-post}
dimred_post     <- TF_fetal_pediatric_epi_post_alve@reductions$umap@cell.embeddings
clustering_post <- TF_fetal_pediatric_epi_post_alve$cell_type2
counts_post_raw <- GetAssayData(TF_fetal_pediatric_epi_post_alve, assay = "RNA", layer = "counts")

sds_post <- load_or_compute(
  ckpt       = CKPT_sds_post,
  save_path  = SAVE_sds_post,
  label      = "slingshot post",
  compute_fn = function() {
    lin <- getLineages(data = dimred_post, clusterLabels = clustering_post,
                       start.clus = TRAJ_START, dist.method = TRAJ_DIST)
    sds <- getCurves(lin, approx_points = TRAJ_APPROX_PTS, thresh = TRAJ_THRESH,
                     stretch = TRAJ_STRETCH, allow.breaks = FALSE, shrink = TRAJ_SHRINK)
    SlingshotDataSet(sds)
  }
)
cat("Post lineages:", length(sds_post@lineages), "\n")
```

```{r plot-slingshot-post, fig.width=FIG_W, fig.height=FIG_H}
Idents(TF_fetal_pediatric_epi_post_alve) <- "cell_type2"
p_col2 <- DimPlot(TF_fetal_pediatric_epi_post_alve, group.by = "cell_type2")
pb2    <- ggplot_build(p_col2)
col_lv2    <- levels(Idents(TF_fetal_pediatric_epi_post_alve))
umap_col2  <- setNames(unique(pb2$data[[1]]$colour), col_lv2)

plot(dimred_post[,1:2], col = umap_col2[as.character(clustering_post)],
     cex = 0.5, pch = 16, xlab = "", ylab = "", axes = FALSE)
lines(sds_post, lwd = 4, col = adjustcolor("steelblue", 0.8))
for (cl in unique(as.character(clustering_post))) {
  idx <- which(as.character(clustering_post) == cl)
  if (length(idx) > 0)
    text(mean(dimred_post[idx,1]), mean(dimred_post[idx,2]), cl, font = 2, cex = 0.9)
}
```

```{r pseudotime-post, fig.width=FIG_W, fig.height=FIG_H}
pt_comb_post <- combined_pt_from_sds(sds_post, rownames(dimred_post))
pt_comb_post[is.na(pt_comb_post)] <- 0
idx_post  <- pmax(1, pmin(200, 1 + floor(scale01(pt_comb_post) * 199)))
pt_cols2  <- pal_pre[idx_post]

plot(dimred_post[,1:2], col = pt_cols2, pch = 16, cex = 0.5, axes = FALSE,
     xlab = "", ylab = "", main = "Post-alve: weighted pseudotime")
for (cl in unique(as.character(clustering_post))) {
  idx <- which(as.character(clustering_post) == cl)
  if (length(idx) > 0)
    text(mean(dimred_post[idx,1]), mean(dimred_post[idx,2]), cl, font = 2, cex = 0.9, col = "darkred")
}
```

---

# 9 · tradeSeq {.tabset}

## 9a · Fit GAMs (pre-merge)

```{r tradeseq-pre}
# Use variable features unless TRADESEQ_GENES is set
genes_pre <- if (!is.null(TRADESEQ_GENES)) TRADESEQ_GENES else
  VariableFeatures(TF_fetal_pediatric_epi_pre_alve)
filt_counts_pre <- counts_pre[genes_pre[genes_pre %in% rownames(counts_pre)], ]

sce_pre <- load_or_compute(
  ckpt       = CKPT_sce_pre,
  save_path  = SAVE_sce_pre,
  label      = "tradeSeq pre",
  compute_fn = function() {
    bp <- MulticoreParam(workers = TRADESEQ_CORES)
    register(bp)
    # sds_pre is the explicit SDS object — fixes the 'curves' variable bug
    sce <- fitGAM(counts = as.matrix(filt_counts_pre), sds = sds_pre,
                  nknots = TRADESEQ_NKNOTS, BPPARAM = bp)
    # Store pseudotime in colData
    pt_gam <- slingPseudotime(sds_pre)
    common  <- intersect(colnames(sce), rownames(pt_gam))
    for (j in seq_len(ncol(pt_gam)))
      colData(sce)[common, paste0("pt_lineage", j)] <- pt_gam[common, j]
    sce
  }
)
plotGeneCount(sds_pre, filt_counts_pre, clusters = clustering_pre, models = sce_pre)
```

```{r tradeseq-tests-pre}
assoRes_pre  <- associationTest(sce_pre)
# Fix: sce_pre passed explicitly (original had 'sce2' undefined)
startRes_pre <- startVsEndTest(sce_pre, lineages = TRUE) %>%
  rownames_to_column("gene") %>% filter(!grepl("^MT-", gene))
patternRes_pre <- patternTest(sce_pre, pairwise = TRUE) %>%
  rownames_to_column("gene") %>% filter(!grepl("^MT-", gene))
earlyDE_pre  <- earlyDETest(sce_pre, knots = c(1, 2))
endRes_pre   <- diffEndTest(sce_pre)

cat("Top startVsEnd genes (pre):\n")
n_lin_pre <- sum(grepl("^waldStat_lineage", colnames(startRes_pre)))
for (k in seq_len(n_lin_pre)) {
  colk <- paste0("waldStat_lineage", k)
  if (colk %in% colnames(startRes_pre)) {
    top5 <- startRes_pre %>% arrange(desc(.data[[colk]])) %>% slice_head(n = 5) %>%
      pull(gene)
    cat("  Lineage", k, ":", paste(top5, collapse = ", "), "\n")
  }
}
```

```{r tradeseq-smoothers-pre, fig.width=10, fig.height=5}
for (g in SMOOTHER_GENES) {
  if (g %in% rownames(sce_pre)) {
    print(plotSmoothers(sce_pre, filt_counts_pre, gene = g) + ggtitle(paste("Pre –", g)))
  }
}
```

## 9b · Fit GAMs (post-merge)

```{r tradeseq-post}
genes_post <- if (!is.null(TRADESEQ_GENES)) TRADESEQ_GENES else
  VariableFeatures(TF_fetal_pediatric_epi_post_alve)
filt_counts_post <- counts_post_raw[genes_post[genes_post %in% rownames(counts_post_raw)], ]

sce_post <- load_or_compute(
  ckpt       = CKPT_sce_post,
  save_path  = SAVE_sce_post,
  label      = "tradeSeq post",
  compute_fn = function() {
    bp <- MulticoreParam(workers = TRADESEQ_CORES)
    register(bp)
    # sds_post passed explicitly
    sce <- fitGAM(counts = as.matrix(filt_counts_post), sds = sds_post,
                  nknots = TRADESEQ_NKNOTS, BPPARAM = bp)
    pt_gam <- slingPseudotime(sds_post)
    common  <- intersect(colnames(sce), rownames(pt_gam))
    for (j in seq_len(ncol(pt_gam)))
      colData(sce)[common, paste0("pt_lineage", j)] <- pt_gam[common, j]
    sce
  }
)
plotGeneCount(sds_post, filt_counts_post, clusters = clustering_post, models = sce_post)
```

```{r tradeseq-tests-post}
assoRes_post   <- associationTest(sce_post)
startRes_post  <- startVsEndTest(sce_post, lineages = TRUE) %>%
  rownames_to_column("gene") %>% filter(!grepl("^MT-", gene))
patternRes_post <- patternTest(sce_post, pairwise = TRUE) %>%
  rownames_to_column("gene") %>% filter(!grepl("^MT-", gene))
earlyDE_post   <- earlyDETest(sce_post, knots = c(1, 2))
endRes_post    <- diffEndTest(sce_post)

cat("Top startVsEnd genes (post):\n")
n_lin_post <- sum(grepl("^waldStat_lineage", colnames(startRes_post)))
for (k in seq_len(n_lin_post)) {
  colk <- paste0("waldStat_lineage", k)
  if (colk %in% colnames(startRes_post)) {
    top5 <- startRes_post %>% arrange(desc(.data[[colk]])) %>% slice_head(n = 5) %>%
      pull(gene)
    cat("  Lineage", k, ":", paste(top5, collapse = ", "), "\n")
  }
}
```

```{r tradeseq-smoothers-post, fig.width=10, fig.height=5}
for (g in SMOOTHER_GENES) {
  if (g %in% rownames(sce_post)) {
    print(plotSmoothers(sce_post, filt_counts_post, gene = g) + ggtitle(paste("Post –", g)))
  }
}
```

## 9c · Heatmaps

```{r tradeseq-heatmaps, fig.width=20, fig.height=8}
for (branch in c("pre","post")) {
  sr   <- if (branch == "pre") startRes_pre else startRes_post
  obj  <- if (branch == "pre") TF_fetal_pediatric_epi_pre_alve else TF_fetal_pediatric_epi_post_alve
  ctcol <- "cell_type3"
  if (!ctcol %in% colnames(obj@meta.data)) ctcol <- "cell_type2"

  topN <- 20
  n_lin <- sum(grepl("^waldStat_lineage", colnames(sr)))
  top_by_lin <- lapply(seq_len(n_lin), function(k) {
    colk <- paste0("waldStat_lineage", k)
    if (!colk %in% colnames(sr)) return(character(0))
    sr %>% arrange(desc(.data[[colk]])) %>% slice_head(n = topN) %>% pull(gene)
  })
  names(top_by_lin) <- paste0("Lineage", seq_len(n_lin))

  avg_exp  <- AverageExpression(obj, group.by = ctcol, assays = "RNA", slot = "data")$RNA
  avg_norm <- setNames(colnames(avg_exp), normalize_labels(colnames(avg_exp)))
  col_fun  <- colorRamp2(c(-2, 0, 2), c("darkgreen","white","red"))

  ht_list <- lapply(seq_len(n_lin), function(k) {
    lin_name    <- paste0("Lineage", k)
    genes_k     <- intersect(top_by_lin[[lin_name]], rownames(avg_exp))
    wanted_disp <- HT_LINEAGE_ORDER[[lin_name]]
    if (is.null(wanted_disp) || length(genes_k) == 0) return(NULL)
    wanted_norm  <- normalize_labels(wanted_disp)
    mapped       <- unname(avg_norm[wanted_norm])
    miss         <- is.na(mapped)
    avg_ext      <- avg_exp
    if (any(miss)) {
      filler <- matrix(NA_real_, nrow(avg_exp), sum(miss),
                       dimnames = list(rownames(avg_exp), wanted_disp[miss]))
      avg_ext <- cbind(avg_exp, filler)
      mapped[miss] <- wanted_disp[miss]
    }
    mat  <- as.matrix(avg_ext[genes_k, mapped, drop = FALSE])
    colnames(mat) <- wanted_disp
    Heatmap(scale_rows(mat), name = paste0("Z(L",k,")"), col = col_fun,
            na_col = "grey95", cluster_rows = FALSE, cluster_columns = FALSE,
            column_title = paste0(branch, " L", k),
            row_title_gp = gpar(fontface = "bold"),
            heatmap_legend_param = list(at = c(-2,0,2), labels = c("Low","Mid","High")))
  })
  ht_list <- Filter(Negate(is.null), ht_list)
  if (length(ht_list) > 0) {
    grid.newpage()
    ht_combined <- Reduce(`+`, ht_list)
    draw(ht_combined, column_title = paste(branch, "branch — top", topN, "genes per lineage"))
  }
}
```

---

# 10 · CellChat {.tabset}

## 10a · Build / load CellChat objects

```{r cellchat-pre}
cellchat_pre <- load_or_compute(
  ckpt       = CKPT_cellchat_pre,
  save_path  = SAVE_cellchat_pre,
  label      = "CellChat pre",
  compute_fn = function() {
    obj  <- TF_fetal_pediatric_epi_pre_alve
    data.input <- GetAssayData(obj, assay = "SCT", layer = "data")
    meta <- data.frame(labels = obj@meta.data[[CC_LABEL_COL]],
                       samples = as.factor(obj@meta.data[[CC_SAMPLE_COL]]),
                       row.names = colnames(obj))
    cc <- createCellChat(object = data.input, meta = meta)
    cc@DB <- CellChatDB.human
    cc <- subsetData(cc)
    future::plan("multisession", workers = 4)
    cc <- identifyOverExpressedGenes(cc)
    cc <- identifyOverExpressedInteractions(cc)
    cc <- smoothData(cc, adj = PPI.human)
    future::plan("sequential")
    options(future.globals.maxSize = 8 * 1024^3)
    cc <- computeCommunProb(cc, type = "triMean", trim = NULL, raw.use = FALSE)
    cc <- filterCommunication(cc, min.cells = CC_MIN_CELLS)
    cc <- computeCommunProbPathway(cc)
    cc <- aggregateNet(cc)
    cc <- netAnalysis_computeCentrality(cc, slot.name = "netP")
    cc
  }
)
cat("Pre CellChat pathways:", length(cellchat_pre@netP$pathways), "\n")
```

```{r cellchat-post}
cellchat_post <- load_or_compute(
  ckpt       = CKPT_cellchat_post,
  save_path  = SAVE_cellchat_post,
  label      = "CellChat post",
  compute_fn = function() {
    obj  <- TF_fetal_pediatric_epi_post_alve
    data.input <- GetAssayData(obj, assay = "SCT", layer = "data")
    meta <- data.frame(labels = obj@meta.data[[CC_LABEL_COL]],
                       samples = as.factor(obj@meta.data[[CC_SAMPLE_COL]]),
                       row.names = colnames(obj))
    cc <- createCellChat(object = data.input, meta = meta)
    cc@DB <- CellChatDB.human
    cc <- subsetData(cc)
    future::plan("multisession", workers = 4)
    cc <- identifyOverExpressedGenes(cc)
    cc <- identifyOverExpressedInteractions(cc)
    cc <- smoothData(cc, adj = PPI.human)
    future::plan("sequential")
    options(future.globals.maxSize = 8 * 1024^3)
    cc <- computeCommunProb(cc, type = "triMean", trim = NULL, raw.use = FALSE)
    cc <- filterCommunication(cc, min.cells = CC_MIN_CELLS)
    cc <- computeCommunProbPathway(cc)
    cc <- aggregateNet(cc)
    cc <- netAnalysis_computeCentrality(cc, slot.name = "netP")
    cc
  }
)
cat("Post CellChat pathways:", length(cellchat_post@netP$pathways), "\n")
```

## 10b · Pathway plots

```{r cellchat-plots, fig.width=8, fig.height=8}
for (pw in CC_PATHWAYS) {
  for (branch in c("pre","post")) {
    cc <- if (branch == "pre") cellchat_pre else cellchat_post
    if (!pw %in% cc@netP$pathways) {
      cat("Pathway", pw, "not found in", branch, "— skipping\n")
      next
    }
    cat("===", pw, "(", branch, ") ===\n")
    netVisual_aggregate(cc, signaling = pw, layout = "circle")
    title(paste(pw, "—", branch, "— circle"), line = 1, cex.main = 1.2)
  }
}
```

```{r cellchat-bubble, fig.width=10, fig.height=6}
for (pw in CC_PATHWAYS) {
  for (branch in c("pre","post")) {
    cc <- if (branch == "pre") cellchat_pre else cellchat_post
    if (!pw %in% cc@netP$pathways) next
    p <- netVisual_bubble(cc,
                           sources.use    = CC_SOURCES,
                           targets.use    = CC_TARGETS,
                           signaling      = pw,
                           remove.isolate = FALSE) +
      ggtitle(paste(pw, "—", branch))
    print(p)
  }
}
```

```{r cellchat-heatmap, fig.width=8, fig.height=6}
for (branch in c("pre","post")) {
  cc <- if (branch == "pre") cellchat_pre else cellchat_post
  for (pw in CC_PATHWAYS) {
    if (pw %in% cc@netP$pathways)
      print(netVisual_heatmap(cc, signaling = pw, color.heatmap = "Reds") +
              ggtitle(paste(pw, "—", branch, "— heatmap")))
  }
}
```

```{r cellchat-signaling-roles, fig.width=10, fig.height=5}
for (branch in c("pre","post")) {
  cc <- if (branch == "pre") cellchat_pre else cellchat_post
  ht1 <- netAnalysis_signalingRole_heatmap(cc, pattern = "outgoing")
  ht2 <- netAnalysis_signalingRole_heatmap(cc, pattern = "incoming")
  draw(ht1 + ht2)
}
```

---

# 11 · Cell Composition {.tabset}

## 11a · Age distributions

```{r composition-age, fig.width=10, fig.height=10}
for (branch in c("pre","post")) {
  obj   <- if (branch == "pre") TF_fetal_pediatric_epi_pre_alve else TF_fetal_pediatric_epi_post_alve
  ctcol <- if ("cell_type3" %in% colnames(obj@meta.data)) "cell_type3" else "cell_type2"
  meta  <- obj@meta.data

  for (cts in list(c("Bud Tip","AT2_2","AT2_1","AT2_3"),
                    c("Bud Tip","AT2_2","Transitional","AT1_5","AT1_4","AT1_3"))) {
    cts <- cts[cts %in% meta[[ctcol]]]
    if (length(cts) == 0) next

    plot_df <- meta[meta[[ctcol]] %in% cts & meta$group %in% c("fetal","pediatric"), ]
    plot_df[[ctcol]] <- factor(plot_df[[ctcol]], levels = cts)
    plot_df$group    <- factor(plot_df$group, levels = c("fetal","pediatric"))

    p <- ggplot(plot_df, aes(x = age_days, fill = group)) +
      geom_histogram(bins = 25, color = "black") +
      facet_grid(reformulate("group", ctcol), scales = "free_x") +
      scale_fill_manual(values = c(fetal = COL_FETAL, pediatric = COL_PED)) +
      theme_classic(base_size = 13) +
      labs(title = paste0(branch, " — age distribution"),
           x = "Age (days)", y = "Cells") +
      guides(fill = "none")
    print(p)
  }
}
```

## 11b · % total per sample

```{r composition-pct, fig.width=10, fig.height=6}
for (branch in c("pre","post")) {
  obj   <- if (branch == "pre") TF_fetal_pediatric_epi_pre_alve else TF_fetal_pediatric_epi_post_alve
  ctcol <- if ("cell_type3" %in% colnames(obj@meta.data)) "cell_type3" else "cell_type2"
  meta  <- obj@meta.data

  total_df <- meta %>% group_by(batch) %>% summarise(total = n(), .groups = "drop")

  pct_all <- meta %>%
    filter(group %in% c("fetal","pediatric")) %>%
    group_by(batch, group, age_days, .data[[ctcol]]) %>%
    summarise(n = n(), .groups = "drop") %>%
    left_join(total_df, by = "batch") %>%
    mutate(pct = 100 * n / total) %>%
    filter(!is.na(age_days), !is.na(pct)) %>%
    arrange(.data[[ctcol]], group, age_days) %>%
    mutate(sample_age = paste0(batch, " (", age_days, ")")) %>%
    group_by(.data[[ctcol]], group) %>%
    mutate(sample_age = factor(sample_age, levels = unique(sample_age))) %>%
    ungroup()

  p <- ggplot(pct_all, aes(x = sample_age, y = pct, fill = group)) +
    geom_col(color = "black", width = 0.85) +
    facet_wrap(reformulate(ctcol), ncol = 2, scales = "free_x") +
    scale_fill_manual(values = c(fetal = COL_FETAL, pediatric = COL_PED)) +
    theme_classic(base_size = 12) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)) +
    labs(title = paste0(branch, " — cell type % of total"),
         x = NULL, y = "% of total cells") +
    guides(fill = "none")
  print(p)
}
```

## 11c · Lineage-filtered composition (Slingshot weights)

```{r composition-lineage, fig.width=10, fig.height=8}
# Uses the pre-merge Slingshot weights
if (exists("sds_pre") && !is.null(sds_pre)) {
  w_sling <- tryCatch(SummarizedExperiment::assay(sds_pre, "weights"),
                       error = function(e) slingCurveWeights(sds_pre))

  lineage_celltypes <- list(
    Lineage1 = c("Bud Tip","AT2_2","Transitional","AT1_5","AT1_4","AT1_1","AT1_2"),
    Lineage2 = c("Bud Tip","AT2_2","Transitional","AT1_5","AT1_4","AT1_3"),
    Lineage3 = c("Bud Tip","AT2_2","AT2_1","AT2_3"),
    Lineage4 = c("Bud Tip","AT2_2","AT2_1","AT2_4")
  )

  meta_pre <- TF_fetal_pediatric_epi_pre_alve@meta.data %>% rownames_to_column("cell")
  ctcol    <- if ("cell_type3" %in% colnames(meta_pre)) "cell_type3" else "cell_type2"

  for (ln in colnames(w_sling)) {
    cts <- lineage_celltypes[[ln]]
    if (is.null(cts)) next
    cells_lin <- rownames(w_sling)[w_sling[, ln] > 0]

    plot_df <- meta_pre %>%
      filter(cell %in% cells_lin, .data[[ctcol]] %in% cts,
             group %in% c("fetal","pediatric")) %>%
      group_by(.data[[ctcol]], group, batch, age_days) %>%
      summarise(n_cells = n(), .groups = "drop") %>%
      arrange(.data[[ctcol]], group, age_days) %>%
      group_by(.data[[ctcol]], group) %>%
      mutate(sample_age = factor(paste0(batch," (",age_days,")"),
                                  levels = unique(paste0(batch," (",age_days,")")))) %>%
      ungroup()
    plot_df[[ctcol]] <- factor(plot_df[[ctcol]], levels = cts)
    plot_df$group    <- factor(plot_df$group, levels = c("fetal","pediatric"))

    p <- ggplot(plot_df, aes(x = sample_age, y = n_cells, fill = group)) +
      geom_col(color = "black", width = 0.85) +
      facet_grid(reformulate("group", ctcol), scales = "free_x", space = "free_x") +
      scale_fill_manual(values = c(fetal = COL_FETAL, pediatric = COL_PED)) +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      labs(title = paste0("Cell counts — ", ln, " (by Slingshot weights)"),
           x = NULL, y = "Cells") +
      guides(fill = "none")
    print(p)
  }
}
```

---

# 12 · Session Info

```{r session-info}
sessionInfo()
```
