---
title: "Fetal–Pediatric Lung snRNA-seq: Final Analysis Report"
subtitle: "Alveolar Lineage Development"
author: "Spence Lab"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 4
    toc-float: true
    theme: flatly
    number-sections: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(targets)
library(yaml)
library(dplyr)
library(ggplot2)
library(knitr)
library(patchwork)

# Load config
paths   <- tar_read(paths)
params  <- tar_read(params)
markers <- tar_read(markers)
```

> **Disclaimer:** This is an observational snRNA-seq study. Causal claims cannot
> be made from trajectory or cell–cell communication analyses alone. Biological
> interpretations are working hypotheses and should be validated experimentally.

# Analysis Overview

This pipeline analyzes pre-processed snRNA-seq data from fetal and pediatric human
lung tissue. The primary goal is to characterize developmental trajectories of
alveolar epithelial cells.

**Two parallel epithelial analysis branches are maintained throughout:**

- **Pre-merge (`pre`):** Epithelial cells subset from individual fetal and pediatric
  objects before joint integration. This approach isolates contaminants per dataset
  before merging.
- **Post-merge (`post`):** Epithelial cells subset from the jointly integrated
  fetal–pediatric object.

Results from both branches are shown side-by-side where applicable.

## Parameters Used

```{r params-summary}
knitr::kable(
  data.frame(
    Parameter = c("PCA dimensions", "Clustering resolution",
                  "Integration method", "Normalization",
                  "Trajectory start", "tradeSeq nknots",
                  "Memory limit (GB)"),
    Value     = c(params$seurat$dims, params$seurat$resolution,
                  params$seurat$integration_method,
                  params$seurat$normalization_method,
                  params$trajectory$start_cluster,
                  params$tradeseq$nknots,
                  params$memory$future_globals_max_size_gb)
  )
)
```

---

# Input Data Summary

```{r input-audit}
audit_dir <- paths$output$audit

load_audit_safe <- function(name, audit_dir) {
  f <- file.path(audit_dir, paste0(name, "_audit.yaml"))
  if (!file.exists(f)) return(NULL)
  yaml::read_yaml(f)
}

ia <- load_audit_safe("input", audit_dir)
if (!is.null(ia)) {
  knitr::kable(data.frame(
    Metric = c("Fetal cells", "Fetal genes", "Fetal batches",
               "Pediatric cells", "Pediatric genes", "Pediatric batches"),
    Value  = c(ia$fetal_n_cells, ia$fetal_n_genes, ia$fetal_batches,
               ia$ped_n_cells,   ia$ped_n_genes,   ia$ped_batches)
  ))
} else {
  cat("Input audit not yet available — run the pipeline first.\n")
}
```

---

# Individual Processing Summary

```{r individual-audit}
iia <- load_audit_safe("individual_integration", audit_dir)
if (!is.null(iia)) {
  knitr::kable(data.frame(
    Metric = c("Fetal cells (post-integration)", "Fetal reductions",
               "Pediatric cells (post-integration)", "Pediatric reductions",
               "Dims used", "Resolution"),
    Value  = c(iia$fetal_n_cells, iia$fetal_reductions,
               iia$ped_n_cells,   iia$ped_reductions,
               iia$dims_used,     iia$resolution)
  ))
} else {
  cat("Individual integration audit not yet available.\n")
}
```

---

# Joint Integration Summary

```{r joint-audit}
ja <- load_audit_safe("joint_integration", audit_dir)
if (!is.null(ja)) {
  knitr::kable(data.frame(
    Metric = c("Total cells (joint)", "Cell types detected", "Lineages"),
    Value  = c(ja$n_cells, ja$cell_types, ja$lineages)
  ))
} else {
  cat("Joint integration audit not yet available.\n")
}
```

---

# Epithelial Branch Comparison

```{r epi-audit}
ea <- load_audit_safe("epithelial_branch", audit_dir)
if (!is.null(ea)) {
  knitr::kable(data.frame(
    Branch     = c("Pre-merge", "Post-merge"),
    `N cells`  = c(ea$pre_n_cells,  ea$post_n_cells),
    `Cell types` = c(ea$pre_cell_types, ea$post_cell_types),
    check.names = FALSE
  ))
} else {
  cat("Epithelial branch audit not yet available.\n")
}
```

**Interpretation note:** Pre-merge subsetting applies contaminant gene filters
within each dataset before integration. Post-merge subsetting relies on the
joint lineage annotation. Differences in cell counts and type composition
between branches reflect this methodological choice, not biological differences.

---

# Alveolar Subset Summary

```{r alve-audit}
aa <- load_audit_safe("alveolar", audit_dir)
if (!is.null(aa)) {
  knitr::kable(data.frame(
    Branch      = c("Pre-alveolar", "Post-alveolar"),
    `N cells`   = c(aa$pre_n_cells,  aa$post_n_cells),
    `Cell types` = c(aa$pre_cell_types, aa$post_cell_types),
    check.names = FALSE
  ))
} else {
  cat("Alveolar audit not yet available.\n")
}
```

---

# Key Figures

## Pre-merge Alveolar UMAP

```{r pre-umap, fig.width=12, fig.height=5}
pre_umap_path <- file.path(paths$output$plots, "alveolar", "pre_alveolar_umap_panel.png")
if (file.exists(pre_umap_path)) {
  knitr::include_graphics(pre_umap_path)
} else {
  cat("UMAP not yet generated — run the pipeline first.\n")
}
```

## Post-merge Alveolar UMAP

```{r post-umap, fig.width=12, fig.height=5}
post_umap_path <- file.path(paths$output$plots, "alveolar", "post_alveolar_umap_panel.png")
if (file.exists(post_umap_path)) {
  knitr::include_graphics(post_umap_path)
} else {
  cat("UMAP not yet generated — run the pipeline first.\n")
}
```

---

# Slingshot Trajectory Results

```{r trajectory-audit}
ta <- load_audit_safe("trajectory", audit_dir)
if (!is.null(ta)) {
  knitr::kable(data.frame(
    Branch         = c("Pre", "Post"),
    `N lineages`   = c(ta$pre_n_lineages,     ta$post_n_lineages),
    `Lineage names` = c(ta$pre_lineage_names,  ta$post_lineage_names),
    check.names = FALSE
  ))
} else {
  cat("Trajectory audit not yet available.\n")
}
```

**Methods note:** Slingshot trajectory inference starts from the Bud Tip
progenitor cluster using minimum-spanning-tree lineage inference. Multiple
start nodes and distance methods are explored. Pseudotime values are
weighted averages across lineage curves.

---

# tradeSeq Pseudotime DE Results

```{r tradeseq-audit}
tsa <- load_audit_safe("tradeseq", audit_dir)
if (!is.null(tsa)) {
  knitr::kable(data.frame(
    Metric  = c("Pre genes tested", "Post genes tested", "nknots", "Cores"),
    Value   = c(tsa$pre_n_genes, tsa$post_n_genes, tsa$nknots, tsa$cores)
  ))
  cat("\n**Fix note:**", tsa$fix_note, "\n")
} else {
  cat("tradeSeq audit not yet available.\n")
}
```

**Methods note:** GAMs were fit using variable features from the alveolar object.
Tests include: association (any pseudotime association), startVsEnd (differential
expression between start and end), pattern (pairwise lineage comparison), earlyDE
(differential expression at early pseudotime), and diffEnd (differential end-point
expression).

```{r tradeseq-tables}
tbl_dir <- file.path(paths$output$tables, "tradeseq")
for (branch in c("pre", "post")) {
  f <- file.path(tbl_dir, paste0(branch, "_tradeseq_startvsend_results.csv"))
  if (file.exists(f)) {
    res <- read.csv(f)
    cat("### ", branch, "-branch: Top 20 startVsEnd genes\n\n")
    print(knitr::kable(head(res[order(res$waldStat, decreasing = TRUE), ], 20)))
    cat("\n")
  }
}
```

---

# CellChat Results

```{r cellchat-audit}
cca <- load_audit_safe("cellchat", audit_dir)
if (!is.null(cca)) {
  knitr::kable(data.frame(
    Branch    = c("Pre", "Post"),
    Pathways  = c(cca$pre_pathways, cca$post_pathways),
    check.names = FALSE
  ))
} else {
  cat("CellChat audit not yet available.\n")
}
```

**Pathways of interest:** BMP, TGFb, FGF — examined for signaling from Bud Tip
and Transitional cells to AT2 subtypes.

---

# Cell Composition Analysis

```{r composition-audit}
compa <- load_audit_safe("composition", audit_dir)
if (!is.null(compa)) {
  knitr::kable(data.frame(
    Branch   = c("Pre-alveolar", "Post-alveolar"),
    `N cells` = c(compa$pre_n_cells, compa$post_n_cells),
    check.names = FALSE
  ))
} else {
  cat("Composition audit not yet available.\n")
}
```

---

# Audit Summary

```{r audit-summary}
source("R/12_audit.R")
all_audits <- consolidate_audits(paths$output$audit)
cat("Total audit records:", length(all_audits), "\n")
cat("Modules audited:", paste(names(all_audits), collapse = ", "), "\n")
```

Full audit details are available in `outputs/audit/`. See `module_audit_report.html`
for a structured per-module breakdown.

---

# Session Information

```{r session-info}
sessionInfo()
```
